Architectures for workflow integration in science gateways

### 2016-04-19 paper.tex
Wrote section 5

Comparison between architectures, per criterion (robustness,
etc). Which architecture is best in each category?

Overall comparison, see Table~\ref{table:overall}.

\begin{table*}
\centering
\begin{tabular}{cccccc}
  \textbf{Tight}
  & \textbf{Service}
  & \textbf{Sub-task}
  & \textbf{Pool}
  & \textbf{Nested}
  & \textbf{Import} \\
\cellcolor[HTML]{99EE99}2.1
  & \cellcolor[HTML]{99EE99}2.1
  & \cellcolor[HTML]{99FE99}1.5
  & \cellcolor[HTML]{99F499}1.9
  & \cellcolor[HTML]{99BB99}4.0
  & \cellcolor[HTML]{99FF99}1.5\\
\end{tabular}
\caption{Overall evaluation. Brighter colors indicate better performance.}
\label{table:overall}
\end{table*}

Limitations: 
\begin{itemize}
\item This ignores pre-existing work (e.g. a workflow engine is already available as a service). Migration across architectures too.
\item The difficulty of implementing interaction g is not properly
reflected. For instance, fsl uses bash, which would be very difficult to convert.
\item We are well aware that all the
components and interactions may not require the same level of
development effort depending on the particular targeted platform.
\end{itemize}

These architectures can be combined (give examples).

>>>>

\subsection{Comparison between architectures}

Tight integration and sub-tasking require the least amount of
integration effort, closely followed by service integration and
workflow import (with service invocation). Pool is a bit behind but
still before nested workflows (with service invocation) which requires
more integration than the other architectures. Robustness leads to the
same ordering of architectures, with tight integration and sub-tasking
in the top group, service integration and workflow import (with
service invocation) close behind, pool in a third group, and nested
workflows (with service invocation) at the end. This ordering is
consistent across metrics, and it reflects the global complexity of
the architectures.

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Sub-tasking and workflow import (with service invocation)
perform a bit better to integrate new engines (\texttt{E$_1$}), which
is interesting given the current profusion of engines with different
characteristics.  However, sub-tasking, workflow import (with service
invocation) and tight integration perform worse than the others to add
engine versions (\texttt{E$_2$}) due to the need to update
infrastructure (sub-tasking), science gateway (tight integration), or
workflow converter (workflow import). Workflow import performs worse
than the others to integrate new workflows (\texttt{E$_3$}) because of
the language conversion step. All architectures except nested
workflows perform the same to integrate new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not a
surprise since it was designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same. The overall scalability score, however, should not
mask the unique characteristics of architectures regarding this
criterion. For instance, nested workflows are the only architecture
that can easily accommodate distributed workflow execution, which can
be critical in some cases. At the same time, the scheduling
constraints created by the sub-tasking and nested workflow
architectures could well become showstoppers depending on the type of
targeted infrastructure. Non-reliable infrastructures, for example,
could hardly cope with workflow engines being wrapped in computing
tasks as done in sub-tasking. The availability of multiple engine
instances could also become a critical feature for science gateways
with important workloads, which would favor sub-tasking and pool.

Differences in other features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. The availability of
fine-grained debugging information may also be critical to efficient
user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in
Table~\ref{table:evaluation}. While Table~\ref{table:overall} provides
a global comparison between architectures, it should not be used
without the detailed metrics reported on Table~\ref{table:evaluation}
and discussed above. Overall, sub-tasking, and workflow import (with
service invocation) perform a bit better than tight integration,
service invocation and pool. Nested workflows stands a bit behind for
the reasons mentioned previously.
\begin{table*}
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Sub-task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Import} \\
  Integration effort (normalized total) & 
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99EF99}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E099}{0.44}
                                    & \cellcolor[HTML]{99BB99}{1.00}
                                      & \cellcolor[HTML]{99E899}{0.33}\\                           
Robustness (normalized total) & 
                                \cellcolor[HTML]{99F599}{0.14}
                                    & \cellcolor[HTML]{99E199}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99CE99}{0.71}
                                    & \cellcolor[HTML]{99BB99}{1.00}
                                    & \cellcolor[HTML]{99E199}{0.43}\\
  Extensibility (normalized total)  & \cellcolor[HTML]{99E099}{0.44}
                                     & \cellcolor[HTML]{99E099}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99EF99}{0.22}
                                     & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99EF99}{0.22}\\
Scalability (normalized total)  & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}\\
Other features (normalized total) & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
                                    & \cellcolor[HTML]{99F499}\textbf{1.9}
                                    & \cellcolor[HTML]{99BB99}\textbf{4.0}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized total scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores
  as done in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

\subsection{Limitations}

A few limitations remain with our evaluation. First of all, the
evaluation is done at a quite high-level of abstraction that may be a
bit artificial when applied to real systems. Abstracting architectures
from their implementation in specific systems is fundamental to the
progress of software engineering of science gateways, but we also
realize that this comes at the cost of realism.

In particular, the presented architectures are pure patterns that may
be blended in actual systems. The distinction between tight
integration and service invocation, for instance, may not be that
clear in practice. Service invocation may also be combined with
sub-tasking in some cases.

Moreover, all the interactions involved in the architectures were
treated equally, as if they all required similar amounts of
development effort, which is realistically not the case. For example,
interaction \texttt{g} (workflow language conversion) clearly requires
more effort than interaction \texttt{d} (service invocation), and may
create much more reliability issues as well. However, quantifying the
robustness and amount of development associated with each interaction
and software component cannot realistically be done without entering
into the details of a particular system.

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed since it is very likely that
this interaction could not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is very tricky because these engines rely
on general-purpose programming languages (bash, Octave, Python, etc)
which are much richer than scientific workflow languages.

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with sub-tasking, could
also be envisaged. We chose to limit ourselves to instantiations with
service invocation because the resulting architectures are implemented in 
real systems, and because service invocation is a largely
used architecture. Nested workflows with sub-tasking, in particular,
could be an interesting architecture to explore.

Finally, there is no need to say that the particular technical or
historical context of a science gateway project may influence the
choice of an architecture to integrate workflow engines. For instance,
some workflow engines may be already available as web services, which
would tend to favor service invocation, and other science gateways may
have strongly tested task and data control features (interactions
\texttt{b} and \texttt{c}), which would favor sub-tasking. The
migration cost between architectures is totally ignored as well.
